---
title: "Expedition Metadata and Sites"
date: today
format: 
  html:
    theme: minty
    self-contained: true
    code-fold: true
    toc: true 
    toc-depth: 3
    toc-location: right
execute:
  fig-width: 10
---

```{r setup, message = F, warning = F, fig.width = 10, fig.height = 10, echo = F}
options(scipen = 999)

library(PristineSeasR2)
library(sf)
library(hms)
library(readxl)
library(janitor)
library(lubridate)
library(gt)
library(pointblank)
library(tidyverse)

ps_paths <- PristineSeasR2::get_drive_paths()

exp_id <- "PNG_2024"

exp_path <- file.path(ps_paths$expeditions, str_replace(exp_id, "_", "-"))

bigrquery::bq_auth(email = "marine.data.science@ngs.org")

bq_connection <- DBI::dbConnect(bigrquery::bigquery(), project = "pristine-seas")
```

```{r}
# Core site field

core_site_fields <- c("ps_site_id","exp_id", "method", "region", "subregion", "locality", "date", "time", "latitude", "longitude", "team_lead", "notes")

# QAQC Functions ----

check_site_duplicates <- function(df, method_name) {
  dups <- df |> 
    count(ps_site_id, sort = TRUE) |> 
    filter(n > 1)
  
  if(nrow(dups) > 0) {
    cat("⚠️ ", method_name, ": Found", nrow(dups), "duplicate site IDs\n")
    print(dups)
  } else {
    cat("✅ ", method_name, ": No duplicate site IDs\n")
  }
  return(dups)
}

check_coordinates <- function(df, method_name, lat_range = c(-40, 40), lon_range = c(-180, 180)) {
  coord_issues <- df |> 
    filter(is.na(latitude) | is.na(longitude) | 
           latitude < lat_range[1] | latitude > lat_range[2] |
           longitude < lon_range[1] | longitude > lon_range[2]) |> 
    select(ps_site_id, latitude, longitude)
  
  if(nrow(coord_issues) > 0) {
    cat("⚠️ ", method_name, ": Found", nrow(coord_issues), "coordinate issues\n")
    print(coord_issues)
  } else {
    cat("✅ ", method_name, ": All coordinates valid\n")
  }
  return(coord_issues)
}

check_dates <- function(df, method_name) {
  date_issues <- df |> 
    filter(is.na(date)) |> 
    select(ps_site_id, date)
  
  if(nrow(date_issues) > 0) {
    cat("⚠️ ", method_name, ": Found", nrow(date_issues), "missing dates\n")
    print(date_issues)
  } else {
    cat("✅ ", method_name, ": All dates present\n")
  }
  return(date_issues)
}

run_qaqc <- function(df, method_name) {
  cat("\n--- QAQC for", method_name, "---\n")
  cat("Total records:", nrow(df), "\n")
  cat("Unique sites:", n_distinct(df$ps_site_id), "\n")
  
  # Run checks
  dups <- check_site_duplicates(df, method_name)
  coords <- check_coordinates(df, method_name)
  dates <- check_dates(df, method_name)
  
  # Return summary
  list(
    method = method_name,
    total_records = nrow(df),
    unique_sites = n_distinct(df$ps_site_id),
    duplicate_sites = nrow(dups),
    coordinate_issues = nrow(coords),
    date_issues = nrow(dates)
  )
}

# Other useful functions ----

## Assign depth strata 
assign_depth_stratum <- function(depth_m, 
                                 shallow_threshold = 14,
                                 super_shallow_threshold = 6) {
  case_when(depth_m == 0 ~ "surface",
            depth_m <= super_shallow_threshold ~ "supershallow",
            depth_m <= shallow_threshold ~ "shallow",
            depth_m <= 30 ~ "deep",
            TRUE ~ NA_character_)}

## Map depth labels (supershallow) to strata "05m"
stratum_to_suffix <- c("surface"      = "00m",
                       "supershallow" = "05m",
                       "shallow"      = "10m",
                       "deep"         = "20m")


# Initialize containers for all methods

all_sites <- list()
qaqc_results <- list()
```

```{r}
region_spelling_fixes <- c("Metenmin" = "Metemin",
                           "Metovai" = "Metevoi",
                           "Tsoi" = "Tsoilik",
                           "Tsolik" = "Tsoilik")
```

## Overview

This script consolidates **site-level metadata** from the **`r exp_id`** expedition fieldbooks into a unified pipeline for validation, visualization, and upload to BigQuery. 

**Workflow**: Import → Standardize → Validate → Visualize → Upload

Site metadata can be processed immediately after fieldwork, providing a foundation for future data integration while downstream analyses (e.g., video processing) are completed.

**Automated QAQC checks:**
- Unique site IDs and valid coordinates  
- Complete dates and spatial coverage
- Interactive maps and summary statistics

### Underwater Visual Surveys

#### QAQC

```{r uvs_sites}
uvs_tbl_colnames <- c(core_site_fields, 
                      "site_name", "habitat", "exposure", "protected", 
                      "blt", "lpi", "inverts", "recruits", "ysi", "edna", "photomosaic")

uvs_habitat_vocab <- c("fore reef", "back reef", "fringing reef", "patch reef", "reef flat","channel","seagrass", "rocky reef")

uvs_exposure_vocab <- c("leeward", "windward", "lagoon")

habitat_lookup <- c("forereef" = "fore reef",
                    "patch" = "patch reef",
                    "passage" = "channel")

uvs_sites <- read_xlsx(file.path(exp_path, "data/primary/raw/_PNG-2024-uvs-metadata.xlsx")) |> 
  clean_names() |> 
  transmute(exp_id = exp_id,
            leg = paste("Leg", leg),
            survey_type = "uvs",
            ps_site_id = str_replace(ps_site_id, "(\\d+)$", ~ str_pad(.x, 3, pad = "0")),
            locality = str_to_title(sublocation), 
            subregion = str_to_title(location),
            region = str_to_title(region),
            date = dmy(dd_mm_yyyy),
            time = as_hms(local_time),
            latitude = as.numeric(lat),
            longitude = as.numeric(lon),
            lead = team_lead,
            notes = NA_character_,
            site_name = NA_character_,
            habitat = str_to_lower(habitat),
            exposure = str_to_lower(exposure),
            protected = as.logical(NA),
            blt = as.logical(NA),
            lpi = as.logical(NA),
            inverts = as.logical(NA),
            recruits = as.logical(NA),
            ysi = as.logical(NA),
            edna = as.logical(NA),
            photomosaic = as.logical(NA)) |> 
  mutate(habitat = recode(habitat, !!!habitat_lookup),
           habitat  = if_else(is.na(habitat), "unknown", habitat),
           exposure = if_else(is.na(exposure), "unknown", exposure)) |> 
  select(all_of(uvs_tbl_colnames))
```

```{r uvs_qaqc, results = "asis"}
validate_uvs_sites <- function(df,
                               habitat_vocab = uvs_habitat_vocab,
                               exposure_vocab = uvs_exposure_vocab,
                               lat_bounds = c(-40, 40),
                               lon_bounds = c(-180, 180),
                               date_bounds = c(as.Date("2024-01-01"), as.Date("2024-12-31"))) {

  fail_list <- list()

  # Check 1: Invalid habitat values
  fail_list$habitat <- df |> 
    filter(!is.na(habitat), !habitat %in% habitat_vocab) |>
    mutate(issue = "Invalid habitat")

  # Check 2: Invalid exposure values
  fail_list$exposure <- df |> 
    filter(!is.na(exposure), !exposure %in% exposure_vocab) |>
    mutate(issue = "Invalid exposure")

  # Check 3: Missing or out-of-bounds coordinates
  fail_list$coords <- df |> 
    filter(is.na(latitude) | is.na(longitude) |
             latitude < lat_bounds[1] | latitude > lat_bounds[2] |
             longitude < lon_bounds[1] | longitude > lon_bounds[2]) |>
    mutate(issue = "Missing or invalid coordinates")

  # Check 4: Duplicate ps_site_id
  fail_list$duplicates <- df |> 
    group_by(ps_site_id) |> 
    filter(n() > 1) |> 
    ungroup() |> 
    mutate(issue = "Duplicate ps_site_id")

  # Check 5: Invalid or out-of-range dates
  fail_list$dates <- df |> 
    filter(is.na(date) |
           date < date_bounds[1] |
           date > date_bounds[2]) |>
    mutate(issue = "Invalid or out-of-range date")

  #  Combine all failures 
  fails_combined <- bind_rows(fail_list) |>
    distinct(ps_site_id, issue, .keep_all = TRUE) |>
    select(ps_site_id, region, subregion, date, habitat, exposure, latitude, longitude, issue)

  if (nrow(fails_combined) == 0) {
    cat("🎉 All UVS site QA/QC checks passed!\n")
  } else {
    message("⚠️ Issues found in UVS site data:")
    print(
      fails_combined |>
        arrange(issue, ps_site_id) |>
        gt() |>
        tab_header(title = "⚠️ UVS Sites QA/QC Issues Summary") |>
        tab_options(table.font.size = "small")
    )
  }
}

validate_uvs_sites(uvs_sites, 
                   date_bounds = c(as.Date("2024-01-01"), as.Date("2024-12-31")))
```

```{r fig-uvs-missing}
#| fig-cap: "Missing data in UVS sites fieldbook"

naniar::vis_miss(uvs_sites)
```

#### Summary

We conducted underwater visual surveys in a total of **`r nrow(uvs_sites)`** sites across **`r n_distinct(uvs_sites$region)`** regions and **`r n_distinct(uvs_sites$subregion)`** subregions of Papua New Guinea. We surveyed a range of habitat and exposures (@tbl-uvs-sites-by-habitat)

```{r}
#| label: tbl-uvs-sites-by-habitat
#| tbl-cap: "Summary of UVS survey effort by region, exposure, and habitat."

uvs_sites |> 
  group_by(leg, region, habitat, exposure) |>
  summarize(n_surveys = n(),
            .groups = "drop") |> 
  pivot_wider(names_from = habitat, values_from = n_surveys, values_fill = 0) |> 
  gt(groupname_col = "leg") |> 
  tab_options(row_group.as_column = T,
              table.width = pct(80),
              table.font.size = 12,
              table.border.top.style = "solid",
              table.border.top.width = px(1),
              table.border.top.color = "black",
              heading.align = "center",
              column_labels.font.weight = "bold")
```

::: {.panel-tabset}

##### By Habitat

```{r}
# Habitat map
habitat_pal <- c("fore reef"  = "#C5FFFD",
                 "back reef"  = "#DF2935",
                 "channel"    = "#D8CC34",
                 "patch reef" = "#772D8B")

uvs_sites_sf <- uvs_sites |>
  mutate(habitat = factor(habitat, levels = names(habitat_pal))) |>
  sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  distinct(ps_site_id, habitat, exposure, geometry)

mapview::mapview(
  uvs_sites_sf,
  zcol = "habitat",
  legend = TRUE,
  col.regions = habitat_pal,
  map.types = "Esri.WorldImagery",
  layer.name = "Habitat",
  popup = leafpop::popupTable(uvs_sites_sf, zcol = c("ps_site_id", "habitat", "exposure"))
)
```

##### By Exposure

```{r}
# Exposure map
exposure_pal <- c("leeward"  = "#1f78b4",
                  "windward" = "#33a02c",
                  "lagoon"   = "#ff7f00")

uvs_sites_sf <- uvs_sites_sf |> 
  mutate(exposure = factor(exposure, levels = names(exposure_pal)))

mapview::mapview(
  uvs_sites_sf,
  zcol = "exposure",
  legend = TRUE,
  col.regions = exposure_pal,
  map.types = "Esri.WorldImagery",
  layer.name = "Exposure",
  popup = leafpop::popupTable(uvs_sites_sf, zcol = c("ps_site_id", "habitat", "exposure"))
)
```

:::

#### Stations

##### Fish - BLT

###### QAQC

```{r eval = T, include = F}
get_fish_stations <- function(file_path,
                              exp_id,
                              sheet = "observations",
                              stratum_to_suffix = c("surface" = "00m",
                                                    "supershallow" = "05m",
                                                    "shallow" = "10m",
                                                    "deep" = "20m")) {
  # Extract expedition ID from file name
  file_name <- basename(file_path)
  full_path <- file.path(exp_path, file_path)
  
  # Load and clean raw data
  raw <- readxl::read_xlsx(full_path,
                           sheet = sheet) |>
    clean_names() |>
    rename(ps_site_id = ps_station_id,
           depth_label = depth_strata) |>
    filter(!is.na(ps_site_id)) |>
    distinct(ps_site_id, depth_label, depth_m, transect, diver) |>
    mutate(ps_site_id = str_replace(ps_site_id, "(\\d+)$", ~ str_pad(.x, 3, pad = "0")),
           depth_m = as.numeric(depth_m),
           depth_label = str_to_lower(depth_label))
  
  # Check for missing depths
   if (any(is.na(raw$depth_m))) {
    warning("⚠️ Missing or non-numeric depths in: ", file_name)
  }
  
  # Summarize and assign strata
  stations <- raw |> 
    group_by(ps_site_id, depth_label, diver) |>
    summarise(depth_m = round(mean(depth_m), 0),
              n_transects = n_distinct(transect),
              area_m2 = n_transects * (25 * 4 + 25 * 2),
              .groups = "drop") |>
    mutate(depth_strata = assign_depth_stratum(depth_m))
  
  # Warn on mismatches
  mismatches <- stations |> 
    filter(depth_label != depth_strata)
  
  if (nrow(mismatches) > 0) {
    message("⚠️ ", nrow(mismatches), " station(s) in ", file_name, " have inconsistent depth labels.")
  }
  
  # Final clean-up and station ID construction
  stations_clean <- stations |>
    group_by(ps_site_id, depth_strata, diver) |>
    summarise(depth_m = round(mean(depth_m), 0),
              n_transects = sum(n_transects),
              area_m2 = sum(area_m2),
              .groups = "drop") |>
    mutate(station_suffix = recode(depth_strata, !!!stratum_to_suffix),
           ps_station_id = paste0(ps_site_id, "_", station_suffix),
           exp_id = exp_id,
           survey_type = "uvs",
           method = "fish") |>
    select(exp_id, survey_type, ps_site_id, ps_station_id, method, everything(), -station_suffix)
  
  # Warn on transect count
  
  flagged <- stations_clean |> 
    filter(n_transects < 2 | n_transects > 3)
  
  if (nrow(flagged) > 0) {
    message("⚠️ ", nrow(flagged), " station(s) less or more transects than expected in: ", file_name)
  }
  
  return(stations_clean)
}
```

```{r fish_stations, results = "asis"}
diver_files <- list(AMF = "PNG_2024_fish_fieldbook_AMF.xlsx",
                    ALG = "PNG_2024_fish_fieldbook_ALG_FINAL.xlsx",
                    VAS = "PNG_2024_fish_fieldbook_VAS.xlsx",
                    JEC = "PNG_2024_fish_fieldbook_JEC_FINAL.xlsx",
                    AJM = "PNG_2024_fish_fieldbook_AJKM_Final_clean.xlsx")

fish_stations <- map_dfr(diver_files, 
                         ~ get_fish_stations(file = file.path("data/primary/raw/fish", .x), 
                                             exp_id = "PNG_2024")) |>
  arrange(ps_station_id) |> 
  left_join(uvs_sites, by = c("exp_id", "ps_site_id", "survey_type") ) |> 
  select(exp_id, survey_type, ps_site_id, ps_station_id, method, diver, depth_m, depth_strata, 
         n_transects, area_m2, date, time, region, subregion, locality, habitat, exposure, notes) |> 
  mutate(exposure = if_else(is.na(exposure), "unknown", exposure),
         habitat = if_else(is.na(habitat), "unknown", habitat))
```

###### Summary

We conducted a total of **`r nrow(fish_stations)`** fish stations (site x depth strata) across **`r n_distinct(fish_stations$subregion)`** locations. We conducted fish BLT surveys at **`r 100*n_distinct(fish_stations$ps_site_id)/n_distinct(uvs_sites$ps_site_id)`** % of the uvs sites and sampled the following habitats and depth strata:

```{r eval = T, include = T}
#| label: tbl-fish-stations
#| tbl-cap: "Summary of fish surveys by habitat and depth strata"

fish_stations |> 
  left_join(uvs_sites |> 
              select(ps_site_id, leg), by = "ps_site_id") |> 
  group_by(leg, region,  habitat, depth_strata) |>
  summarise(n = n_distinct(ps_station_id), .groups = "drop") |> 
  pivot_wider(names_from = habitat, values_from = n, values_fill = 0) |> 
  gt(groupname_col = "leg") |> 
  tab_options(table.font.size = 12,
              table.width = pct(80),
              row_group.as_column = T) |> 
  tab_source_note(source_note = "Depth stratum: supershallow (< 5 m), shallow (5 - 15 m), and deep (>= 15 m).")
```

##### Benthos - LPI

###### QAQC

```{r eval = T, include = F}
get_lpi_stations <- function(file_path,
                             sheet,
                             range,
                             exp_id,
                             diver,
                             stratum_to_suffix = c("surface" = "00m",
                                                   "supershallow" = "05m",
                                                   "shallow" = "10m",
                                                   "deep" = "20m"),
                             min_sections = 5){
  
  file_name <- basename(file_path)
  full_path <- file.path(exp_path, file_path)
  
  # Step 1: Read and clean raw LPI observations
  df_raw <- readxl::read_xlsx(path = full_path,
                              sheet = sheet,
                              range = range,
                              col_names = FALSE,
                              .name_repair = "minimal") |>
    t() |>
    as_tibble(.name_repair = "minimal") |>
    set_names(c("station", "section", "depth_m")) |>
    mutate(station     = str_to_upper(as.character(station)),
           section     = as.character(section),
           depth_m     = readr::parse_number(depth_m),
           site_num    = str_extract(station, "\\d+") |> str_pad(3, pad = "0"),
           ps_site_id  = paste(exp_id, "uvs", site_num, sep = "_"),
           exp_id      = exp_id,
           survey_type = "uvs",
           method      = "lpi",
           diver       = diver) |>
    filter(!is.na(depth_m))
  
  # Step 2: Summarize by station
  df <- df_raw |>
    group_by(exp_id, survey_type, ps_site_id, station, method, diver) |>
    summarise(depth_m    = round(mean(depth_m, na.rm = TRUE), 0),
              n_sections = n_distinct(section),
              .groups    = "drop") |>
    mutate(length_m       = n_sections * 10,
           depth_strata   = assign_depth_stratum(depth_m),
           station_suffix = recode(depth_strata, !!!stratum_to_suffix),
           ps_station_id  = paste0(ps_site_id, "_", station_suffix)) |>
    select(exp_id, survey_type, ps_site_id, ps_station_id,
           method, diver, depth_m, depth_strata,
           n_sections, length_m)
  
  # Step 3: Flag short transects
  flagged <- df |> 
    filter(n_sections < min_sections)

  if (nrow(flagged) > 0) {
    message("⚠️ ", nrow(flagged), " station(s) have fewer than ", min_sections, " sections in:",file_name)
  }

  return(df)
}
```

```{r eval = T, results = "asis"}
lpi_stations_leg2 <- get_lpi_stations(file  = "data/primary/raw/benthos/PNG_2024_lpi_fieldsheet_Emma.xlsx",
                                      sheet = "transects",
                                      range = "F3:ND5",
                                      exp_id = "PNG_2024",
                                      diver  = "EC")

lpi_stations_leg1 <- get_lpi_stations(file  = "data/primary/raw/benthos/PNG_2024_lpi_fieldsheet_Quim.xlsx",
                                      sheet = "transects",
                                      range = "I3:PL5",
                                      exp_id = "PNG_2024",
                                      diver  = "QG")

lpi_stations_kat <- get_lpi_stations(file  = "data/primary/raw/benthos/PNG_2024_lpi_fieldsheet_Kat.xlsx",
                                     sheet = "transects",
                                     range = "H2:Q4",
                                     exp_id = "PNG_2024",
                                     diver  = "KM")

lpi_stations <- bind_rows(lpi_stations_leg1, 
                          lpi_stations_leg2,
                          lpi_stations_kat) |> 
  select(exp_id, survey_type, method, ps_site_id, ps_station_id, depth_strata, depth_m, n_sections, length_m, diver) |> 
  left_join(uvs_sites, 
            by = c("exp_id", "ps_site_id", "survey_type") ) |> 
  select(exp_id, survey_type, ps_site_id, method, ps_station_id, diver, depth_m, depth_strata, 
         n_sections, length_m, date, time, region, subregion, locality, habitat, exposure, notes) 
```

###### Summary

We conducted a total of **`r nrow(lpi_stations)`** LPI stations (site x depth strata) across **`r n_distinct(lpi_stations$subregion)`** locations. We conducted benthic LPI surveys at **`r round(100*n_distinct(lpi_stations$ps_site_id)/n_distinct(uvs_sites$ps_site_id))`** % of the uvs sites and sampled the following habitats and depth strata:

```{r eval = T, include = T}
#| label: tbl-lpi-stations
#| tbl-cap: "Summary of benthic surveys by habitat and depth strata"

lpi_stations |> 
  left_join(uvs_sites |> 
              select(ps_site_id, leg), by = "ps_site_id") |> 
  group_by(leg, region, habitat, depth_strata) |>
  summarise(n = n_distinct(ps_station_id), .groups = "drop") |> 
  pivot_wider(names_from = habitat, values_from = n, values_fill = 0) |> 
  gt(groupname_col = "leg") |> 
  tab_options(table.font.size = 12,
              table.width = pct(80),
              row_group.as_column = T) |> 
  tab_source_note(source_note = "Depth stratum: supershallow (< 5 m), shallow (5 - 15 m), and deep (>= 15 m).")
```

###### What stations have fish surveys but not LPI?

```{r, results = "asis"}
# 1. Get station-level presence for each method
fish_ids <- fish_stations |> distinct(ps_station_id) |> mutate(has_fish = TRUE)
lpi_ids  <- lpi_stations  |> distinct(ps_station_id) |> mutate(has_lpi = TRUE)

# 2. Full join to capture all stations
station_compare <- full_join(fish_ids, lpi_ids, by = "ps_station_id") |>
  mutate(across(c(has_fish, has_lpi), ~replace_na(.x, FALSE))) |>
  filter(!(has_fish & has_lpi))  # Keep only those missing one

# 3. Optional: add context (e.g., from fish_stations)
station_details <- fish_stations |> 
  bind_rows(lpi_stations) |> 
  distinct(ps_station_id, ps_site_id, depth_strata, exp_id)  # light joinable metadata

mismatch_table <- station_compare |> 
  left_join(station_details, by = "ps_station_id") |>
  select(exp_id, ps_site_id, ps_station_id, has_fish, has_lpi, depth_strata) |>
  arrange(ps_station_id)

# 4. Display
if (nrow(mismatch_table) == 0) {
  cat("✅ All stations have both Fish and LPI surveys.\n")
} else {
  message("⚠️ ", nrow(mismatch_table), " station(s) are missing either fish or LPI:")
  print(
    mismatch_table |> 
      select(ps_station_id, has_fish, has_lpi, depth_strata) |>
      gt() |> 
      tab_header(title = md("⚠️ **Stations Missing One Method (Fish or LPI)**")) |> 
      tab_options(table.font.size = 12)
  )
}
```

##### Coral recruits

###### QAQC

```{r}
recruits_stations <- readxl::read_xlsx(file.path(exp_path, "data/primary/raw/benthos/PNG_2024_recruits_fieldsheet_MT.xlsx"), 
                                       sheet = "Coral Recruits") |> 
  clean_names() |>
  rename(depth_m = depth, n_quadrats = quadrats, diver = observer) |> 
  mutate(exp_id         = exp_id,
         survey_type    = "uvs",
         depth_m        = round(as.numeric(depth_m), 0),
         site_num       = str_extract(site, "\\d+") |> str_pad(3, pad = "0"),
         ps_site_id     = paste(exp_id, "uvs", site_num, sep = "_"),
         depth_strata   = assign_depth_stratum(depth_m),
         station_suffix = recode(depth_strata, !!!stratum_to_suffix),
         ps_station_id  = paste0(ps_site_id, "_", station_suffix),
         method         = "recruits") |> 
  group_by(exp_id, survey_type, ps_site_id, ps_station_id, diver, depth_m, depth_strata, method) |> 
  summarise(n_quadrats = mean(n_quadrats, na.rm = TRUE),
            .groups = "drop") |> 
  select(exp_id, survey_type , ps_site_id, ps_station_id, method, diver, depth_m, depth_strata, n_quadrats) |>
  arrange(ps_station_id) |> 
  left_join(uvs_sites, 
            by = c("exp_id", "survey_type","ps_site_id") ) |> 
  select(exp_id, survey_type, ps_site_id, method, ps_station_id, diver, depth_m, depth_strata, n_quadrats, date, time, 
         region, subregion, locality, habitat, exposure, notes) 
```

###### Summary

We conducted a total of **`r nrow(recruits_stations)`** coral recruits stations (site x depth strata) across **`r n_distinct(recruits_stations$subregion)`** subregions We conducted coral recruits surveys at **`r round(100*n_distinct(recruits_stations$ps_site_id)/n_distinct(uvs_sites$ps_site_id))`** % of the uvs sites.

##### Invertebrates


###### QAQC

```{r}
inverts_stations <- readxl::read_xlsx(file.path(exp_path, 
                                                "data/primary/raw/benthos/PNG_2024_inverts_fieldsheet_MT.xlsx"), 
                                      sheet = "Invert") |> 
  clean_names() |> 
  filter(transect != "Off") |> 
  rename(depth_m       = depth, 
         diver         = observer) |> 
  mutate(exp_id        = exp_id,
         survey_type   = "uvs",
         method        = "inverts",
         depth_m       = round(as.numeric(depth_m), 0),
         site_num      = str_extract(site, "\\d+") |> str_pad(3, pad = "0"),
         ps_site_id    = paste(exp_id, "uvs", site_num, sep = "_"),
         depth_strata  = assign_depth_stratum(depth_m),
         station_suffix = recode(depth_strata, !!!stratum_to_suffix),
         ps_station_id  = paste0(ps_site_id, "_", station_suffix) ) |> 
  distinct(exp_id, survey_type, ps_site_id, ps_station_id, method, diver, depth_m, depth_strata) |> 
  arrange(ps_station_id)

inverts_stations <- inverts_stations |> 
  left_join(uvs_sites, by = c("exp_id", "survey_type","ps_site_id")) |> 
  select(exp_id, survey_type, ps_site_id, method, ps_station_id, diver, depth_m, depth_strata, date, time, 
          region, subregion, locality,  habitat, exposure, notes)
```

###### Summary

We conducted a total of **`r nrow(inverts_stations)`** invertebrate stations (site x depth strata) across **`r n_distinct(inverts_stations$subregion)`** locations. We conducted invertebrate surveys at **`r round(100*n_distinct(inverts_stations$ps_site_id)/n_distinct(uvs_sites$ps_site_id))`** % of the uvs sites.

##### eDNA

###### QAQC

```{r}
edna_samples <- read_xlsx(file.path(exp_path, "data/primary/raw/edna/PNG_2024_edna_fieldbook.xlsx"),
                          sheet = "metadata") |>
  clean_names() |>
  mutate(exp_id         = exp_id, 
         method         = "edna",
         survey_type    = "edna",
         leg            = paste("Leg", leg),
         depth_m        = round(as.numeric(depth_m)),
         replicate      = str_extract(ps_sample_id, "\\d+$"),
         site_num       = str_extract(ps_station_id, "\\d+$") |> str_pad(3, pad = "0"),
         ps_site_id     = paste(exp_id, "edna", site_num, sep = "_"),
         depth_strata   = assign_depth_stratum(depth_m),
         station_suffix = recode(assign_depth_stratum(depth_m), !!!stratum_to_suffix),
         ps_station_id  = paste0(ps_site_id, "_", station_suffix),
         ps_sample_id   = paste0(ps_station_id, "_r", replicate),
         paired_site_id = str_replace(paired_station_id, "\\d+$", \(x) str_pad(x, 3, pad = "0")),
         date           = ymd(date),
         time           = as_hms(collection_time),
         filter_date    = ymd(filter_date),
         filter_time    = as_hms(filter_time))

# === Add region & finalize fields ===
region_lookup <- uvs_sites |> distinct(location = subregion, region)

edna_samples <- edna_samples |>
  left_join(region_lookup, by = "location") |>
  rename(subregion = location, 
         locality = sublocation,
         latitude = lat, 
         longitude = lon, 
         lead = team_lead) |> 
  mutate(region = case_when(subregion %in% c("Metemin", "Metevoi") ~ "Lovongai", 
                            TRUE ~ region)) |> 
  select(all_of(core_site_fields),
         method, ps_station_id, paired_site_id, ps_sample_id, replicate, depth_strata, depth_m, water_liters,
         preservative, filter_date, filter_time, filter_type, filter_size_um, target, habitat, site_photos) |>
  arrange(ps_station_id)

# === Sites summary ===
edna_sites <- edna_samples |>
  group_by(exp_id, leg, survey_type, ps_site_id, region, subregion, locality,
           date, time, latitude, longitude, lead, paired_site_id, habitat, site_photos) |>
  summarize(n_stations = n_distinct(ps_station_id),
            n_samples  = n_distinct(ps_sample_id),
            notes      = first(na.omit(notes)), .groups = "drop")

# === Stations summary ===
edna_stations <- edna_samples |>
  group_by(exp_id, method, ps_station_id, ps_site_id, paired_site_id, depth_strata,
           depth_m, filter_type, filter_size_um, filter_date, filter_time, target) |>
  summarize(n_replicates = n_distinct(ps_sample_id),
            water_liters = sum(water_liters, na.rm = TRUE),
            notes        = first(na.omit(notes)), .groups = "drop") |>
  relocate(ps_site_id, .after = ps_station_id)

# === Flag stations with missing replicates ===
flagged <- edna_stations |> filter(n_replicates != 3)

if (nrow(flagged) > 0) {
  cli::cli_warn("⚠️ {nrow(flagged)} station(s) have fewer or more than 3 eDNA replicates.")
}
```

###### Summary

We collected a total of **`r nrow(edna_samples)`** eDNA samples across **`r n_distinct(edna_sites$subregion)`** locations. We conducted eDNA surveys at **`r round(100*n_distinct(edna_sites$ps_site_id)/n_distinct(uvs_sites$ps_site_id))`** % of the uvs sites.

##### YSI

###### QAQC

```{r warning = FALSE}
ysi_sites_leg_1 <- read_xlsx(file.path(exp_path, "data/primary/raw/ysi/PNG_2024_ysi_fieldbook_leg_1.xlsx")) |>
  mutate(ps_site_id = str_replace(ps_station_id, "(\\d+)$", \(x) str_pad(x, 3, pad = "0"))) |> 
  distinct(ps_site_id) |> 
  pull()

ysi_raw_leg_2 <- list.files(path = file.path(exp_path, "data/primary/raw/ysi"),
                              pattern = "KorDSS Measurement File Export.*\\.csv$",
                              full.names = TRUE) |> 
  read_csv(skip = 5, show_col_types = FALSE) |>
  clean_names() |> 
  mutate(date = mdy(date),
         datetime = as_datetime(date) + time)

ysi_casts_leg_2 <- ysi_raw_leg_2 |> 
  arrange(datetime) |>
  mutate(time_diff_secs = as.numeric(difftime(datetime, lag(datetime), units = "secs")),
         cast_id = cumsum(if_else(is.na(time_diff_secs) | time_diff_secs > 300, 1, 0))) |> 
  group_by(cast_id) |>
  summarise(cast_duration_sec = as.numeric(difftime(max(datetime), min(datetime), units = "secs")),
            datetime = min(datetime),
            n_pings = n(),
            max_depth_m = max(depth_m, na.rm = TRUE),
            .groups = "drop") |> 
  filter(cast_duration_sec >= 30,  # at least 30 seconds long
         n_pings >= 5 )             # at least 5 pings
         
library(fuzzyjoin)

ysi_casts_matched <- fuzzy_left_join(ysi_casts_leg_2,
                                     uvs_sites |> 
                                       mutate(datetime = as_datetime(date) + time) |> 
                                       select(ps_site_id, datetime),
                                     by = c("datetime" = "datetime"),
                                     match_fun = list(\(x, y) abs(difftime(x, y, units = "mins")) < 40))


# After manually reviewed the casts, we will make the manual adjustement

ysi_casts_matched$ps_site_id[ysi_casts_matched$cast_id == 22] <- "PNG_2024_uvs_062"
ysi_casts_matched$ps_site_id[ysi_casts_matched$cast_id == 29] <- "PNG_2024_uvs_061"
ysi_casts_matched$ps_site_id[ysi_casts_matched$cast_id == 48] <- "PNG_2024_uvs_088"

ysi_casts_matched <- ysi_casts_matched |>
  filter(!is.na(ps_site_id)) |> 
  rename(datetime = datetime.x) |> 
  select(-datetime.y) |> 
  mutate(method = "ysi",
         ps_station_id = ps_site_id)

ysi_sites <- tibble(ps_site_id = unique(c(ysi_casts_matched$ps_site_id, ysi_sites_leg_1))) |> 
  left_join(uvs_sites, by = "ps_site_id") |> 
  filter(ps_site_id != "PNG_2024_uvs_SPAG") |> 
  rename(paired_site_id = ps_site_id) |> 
  arrange(paired_site_id) |> 
  mutate(survey_type = "ysi",
         cast_number = str_pad(row_number(), 3, pad = "0"),
         ps_site_id = paste(exp_id, survey_type, cast_number, sep = "_")) |>
  arrange(ps_site_id) |> 
  select(c(core_site_fields, "paired_site_id", "habitat", "exposure"))
```

###### Summary

We conducted a total of **`r nrow(ysi_sites)`** YSI casts across **`r n_distinct(ysi_sites$subregion)`** distinct subregions We conducted YSI surveys at **`r round(100*length(ysi_sites)/n_distinct(uvs_sites$ps_site_id))`** % of the uvs sites.

#### Site coverage

The following table summarizes the coverage of each method across the UVS sites:

```{r}
#| label: fig-uvs-heatmap
#| fig-cap: "Coverage of UVS survey methods (fish, LPI, recruits, inverts, eDNA) at each site."

uvs_sites <- uvs_sites |> 
  mutate(blt         = ps_site_id %in% fish_stations$ps_site_id,
         lpi         = ps_site_id %in% lpi_stations$ps_site_id,
         inverts     = ps_site_id %in% inverts_stations$ps_site_id,
         recruits    = ps_site_id %in% recruits_stations$ps_site_id,
         edna        = ps_site_id %in% edna_sites$paired_site_id,
         ysi         = ps_site_id %in% ysi_sites$paired_site_id,
         photomosaic = FALSE)

uvs_sites |> 
  select(leg, ps_site_id, blt, lpi, inverts, recruits, ysi, edna) |> 
  pivot_longer(-c(leg, ps_site_id), 
               names_to = "method", 
               values_to = "conducted") |> 
  mutate(site_number = str_extract(ps_site_id, "\\d+$")) |>
  ggplot(aes(x = method, y = site_number, fill = conducted)) +
  geom_tile(color = "white", 
            linewidth = 0.1) +
  scale_fill_manual(values = c("TRUE" = "#1a9850", "FALSE" = "#f0f0f0")) +
  labs(title = "Survey Method Coverage by UVS Site",
       x = "", y = "", 
       fill = "Conducted") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  facet_wrap(~leg, scales = "free_y")
```

```{r}
#| label: fig-uvs-upset
#| fig-cap: "Patterns of data overlap across UVS methods (BLT, LPI, inverts, recruits, YSI)."
#| 
library(UpSetR)

method_sets <- uvs_sites |> 
  select(ps_site_id, blt, lpi, inverts, recruits, ysi, edna) |> 
  pivot_longer(-ps_site_id, names_to = "method", values_to = "present") |> 
  filter(present == TRUE) |> 
  group_by(method) |> 
  summarise(sites = list(ps_site_id), .groups = "drop") |> 
  deframe()

# Generate UpSet plot
upset(fromList(method_sets))
```

```{r}
# Sites object with only UVS sites
sites <- list(uvs = uvs_sites,
              edna = edna_sites,
              ysi = ysi_sites)

# Stations object with fish and LPI stations
stations <- list(fish = fish_stations,
                 lpi  = lpi_stations,
                 recuits = recruits_stations,
                 inverts = inverts_stations)

write_rds(sites, 
          file.path(exp_path, "data/primary/processed/sites.rds"))

write_rds(stations, 
          file.path(exp_path, "data/primary/processed/stations.rds"))
```



### Seabed BRUVS

#### QAQC

```{r}
region_lookup <- uvs_sites |> distinct(location = subregion, region)

sbruvs_raw <- read_xlsx(file.path(exp_path, "data/primary/raw/bruvs/PNG_2024_bruvs_fieldbook.xlsx"),
                          sheet = "metadata") |>
  clean_names() |> 
  rename(lead = team_lead) |> 
  mutate(exp_id = "PNG_2024",
         survey_type = "sbruvs",
         method = "seabed BRUVS",
         ps_site_id =   str_replace(ps_station_id, "(\\d+)$", \(x) str_pad(x, 3, pad = "0")),
         ps_station_id = paste(ps_site_id, paste0(round(depth_m), "m"), sep = "_"),
         date = ymd(date),
         time = as_hms(time_in),
         depth_m = as.numeric(depth_m),
         location = str_to_title(location),
         sublocation = str_to_title(sublocation),
         habitat = NA_character_,
         exposure = NA_character_) 
  
sbruvs_stations <- sbruvs_raw |>
  left_join(region_lookup, by = "location") |>
  rename(subregion = location, 
         locality = sublocation) |>
  mutate(region = case_when(subregion == "Emanaus" ~ "Murat", 
                            subregion %in% c("Tsoilik", "Metenmin", "Anelava", "Metevoi") ~ "Lovongai", 
                            TRUE ~ region)) |> 
  select(all_of(core_site_fields),
         ps_station_id, depth_m, rig, cam_l, cam_r, habitat, exposure, bottom_type)

sbruvs_sites <- sbruvs_stations |>
  select(all_of(core_site_fields), habitat, exposure) |> 
  mutate(habitat = if_else(is.na(habitat), "unknown", habitat),
         exposure = if_else(is.na(exposure), "unknown", exposure)) |> 
  distinct()

sbruvs_stations <- sbruvs_stations |> 
  select(exp_id, survey_type, ps_station_id, ps_site_id, depth_m, rig, cam_l, cam_r, bottom_type, notes)
```

#### Summary

We deployed a total of **`r nrow(sbruvs_sites)`** seabed BRUVS distributed across **`r n_distinct(sbruvs_sites$subregion)`** locations and **`r n_distinct(sbruvs_sites$region)`** study regions. 

```{r}
# Regions palette

regions_pal <- c("Lovongai"  = "#FABC2A",
                "Manus"  = "#44FFD1",
                "Murat"    = "#275DAD",
                "Western Islands" = "#FF69EB")

sbruvs_sites_sf <- sbruvs_sites |>
  mutate(region = factor(region, levels = names(regions_pal))) |>
  sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  distinct(ps_site_id, region, exposure, geometry)

mapview::mapview(sbruvs_sites_sf,
                 zcol = "region",
                 legend = TRUE,
                 col.regions = regions_pal,
                 map.types = "Esri.WorldImagery",
                 layer.name = "Region",
                 popup = leafpop::popupTable(sbruvs_sites_sf, zcol = c("ps_site_id", "region", "exposure")))
```


### Pelagic BRUVS

#### QAQC

```{r}
qaqc_warn_pbruvs <- function(df,
                             max_drift_m = 5000,
                             max_drift_hours = 3,
                             expected_n_rigs = 5) {
  require(dplyr)
  
  # Drift distance
  n_drift_exceeds <- df |> filter(drift_m > max_drift_m) |> nrow()
  if (n_drift_exceeds > 0) message(glue::glue("⚠️  {n_drift_exceeds} station(s) with drift > {max_drift_m} m"))

  # Drift time issues
  n_drift_zero <- df |> filter(drift_hrs <= 0) |> nrow()
  if (n_drift_zero > 0) message("⚠️  ", n_drift_zero, " station(s) with nonpositive drift time")

  n_drift_exceeds_time <- df |> filter(drift_hrs > max_drift_hours) |> nrow()
  if (n_drift_exceeds_time > 0) message(glue::glue("⚠️  {n_drift_exceeds_time} station(s) with drift time > {max_drift_hours} hours"))

  # Missing coordinates
  n_missing_coords <- df |> filter(is.na(lat_in) | is.na(long_in) | is.na(lat_out) | is.na(long_out)) |> nrow()
  if (n_missing_coords > 0) message("⚠️  ", n_missing_coords, " station(s) with missing coordinates")

  # Too few rigs per site
  n_few_rigs <- df |> count(ps_site_id) |> filter(n < expected_n_rigs) |> nrow()
  if (n_few_rigs > 0) message("⚠️  ", n_few_rigs, " site(s) with fewer than ", expected_n_rigs, " rigs")

  # Duplicate rigs
  n_dup_rigs <- df |> count(ps_site_id, rig) |> filter(n > 1) |> nrow()
  if (n_dup_rigs > 0) message("⚠️  ", n_dup_rigs, " duplicated rig(s) within site(s)")

  # If no issues found
  if (n_drift_exceeds + n_drift_zero + n_drift_exceeds_time + n_missing_coords + n_few_rigs + n_dup_rigs == 0) {
    message("✅ No QAQC issues found in pBRUV stations.")
  }
}
```

```{r}
pbruvs_raw <- read_xlsx(file.path(exp_path, "data/primary/raw/pelagics/PNG_2024_pelagics_fieldbook_2024_10_01.xlsx"),
                        sheet = "Metadata") |>
  clean_names() 

pbruvs_stations <- pbruvs_raw |> 
  mutate(exp_id = "PNG_2024",
         survey_type = "pbruvs",
         method = "pelagic BRUVS",
         ps_site_id = str_replace(string, "(\\d+)$", \(x) str_pad(x, 3, pad = "0")),
         ps_site_id = str_replace(ps_site_id, "PNGP24", "PNG_2024_pbruvs"),
         ps_station_id = paste(ps_site_id, paste0("r",round(rig) ), sep = "_"),
         sublocation = str_to_title(sublocation),
         location = str_to_title(location),
         date = ymd(date),
         time_in = as_hms(time_in),
         time_out = as_hms(time_out),
         drift_hrs = as.numeric(round(difftime(time_out, time_in, units = "hours"),2)),
         drift_m =  round(as.numeric(geosphere::distHaversine(matrix(c(long_in, lat_in), ncol = 2),
                                                              matrix(c(long_out, lat_out), ncol = 2))), 2))|> 
  rename(uwa_opcode = opcode, uwa_string = string, notes = field_comments) |> 
  select(exp_id, leg, survey_type, method, ps_site_id, ps_station_id, location, sublocation, date, 
         rig, left_cam, right_cam, time_in, time_out, contains("lat_"), contains("long_"), drift_hrs, drift_m,
         bait, uwa_opcode, uwa_string, species_notes, notes, team_lead) |> 
  left_join(region_lookup, by = "location") |>
  rename(subregion = location, 
         locality = sublocation,
         lead = team_lead) |>
  mutate(region = case_when(subregion %in% c("Emanaus", "Eloaua") ~ "Murat", 
                            subregion %in% c("Tsoilik", "Metenmin", "Anelava", "Metovai") ~ "Lovongai", 
                            TRUE ~ region))

# Run QAQC

qaqc_warn_pbruvs(pbruvs_stations)
 
# Sites table

pbruvs_sites <- pbruvs_stations |> 
  group_by(exp_id, leg, survey_type, ps_site_id, uwa_string, region, subregion, locality, date, lead) |> 
  summarise(time = first(na.omit(time_in)),
            latitude = mean(lat_in, na.rm = T),
            longitude = mean(long_in, na.rm = T),
            n_rigs = n_distinct(ps_station_id), 
            drift_m = round(mean(drift_m),2),
            drift_hrs = round(mean(drift_hrs),2),
            notes = first(na.omit(notes)),
            .groups = "drop") |> 
  select(all_of(core_site_fields), n_rigs, drift_m, drift_hrs) 
```

#### Summary

We deployed a total of **`r nrow(pbruvs_sites)`** pelagic BRUVS distributed across **`r n_distinct(pbruvs_sites$subregion)`** locations and **`r n_distinct(pbruvs_sites$region)`** study regions.

```{r}
pbruvs_sites_sf <- pbruvs_sites |>
  sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  distinct(ps_site_id, region, drift_hrs, drift_m, n_rigs, geometry)

mapview::mapview(pbruvs_sites_sf,
                 zcol = "region",
                 legend = TRUE,
                 col.regions = regions_pal,
                 map.types = "Esri.WorldImagery",
                 layer.name = "Region",
                 popup = leafpop::popupTable(pbruvs_sites_sf, zcol = c("ps_site_id", "region", "n_rigs", "drift_hrs","drift_m")))
```

### Birds

#### QAQC

```{r warning = FALSE}
birds_raw <- read_xlsx(file.path(exp_path, "data/primary/raw/birds/PNG_2024_birds_fieldbook_2024_10_02.xlsx"),
                        sheet = "transect",
                       range = cell_cols("A:R")) |>
  clean_names()

birds_stations <- birds_raw |> 
  select(-time) |> 
  rename(leg = exp_leg, 
         uwa_string = paired_string_id,
         lead = team_member) |> 
  mutate(exp_id = "PNG_2024",
         survey_type = "birds",
         method = "transects",
         ps_station_id = ps_site_id,
         date = ymd(date),
         across(c(start_time, stop_time), as_hms),
         across(c(start_lat, start_lon, stop_lat, stop_lon, distance), as.numeric),
         duration_hrs = as.numeric(round(difftime(stop_time, start_time, units = "hours"), 2)),
         linear_dist_m = round(as.numeric(geosphere::distHaversine(matrix(c(start_lon, start_lat), ncol = 2),
                                                                 matrix(c(stop_lon, stop_lat), ncol = 2))), 2),
         distance_m = coalesce(distance*1000, linear_dist_m)) |> 
  select(exp_id, method, ps_site_id, ps_station_id, location , sublocation , date, contains("time"), contains("start_l"), contains("stop_l"), 
         vessel, duration_hrs, distance_m, uwa_string, in_or_out, notes, everything()) 

birds_stations <- birds_stations |> 
  left_join(region_lookup, by = "location") |>
  rename(subregion = location, 
         locality = sublocation) |> 
  mutate(region = case_when(subregion %in% c("Emanaus", "Eloaua") ~ "Murat", 
                            subregion %in% c("Tsoi","Tsoilik", "Metenmin", "Anelava", "Metovai") ~ "Lovongai", 
                            TRUE ~ region)) 

birds_sites <- birds_stations |> 
  mutate(time = start_time, 
         latitude = start_lat,
         longitude = start_lon,
         habitat = case_when(vessel %in% c("Argo", "Rhib") ~ "coastal",
                             vessel %in% c("Land", "Land/bike") ~ "inland",
                             TRUE ~ NA_character_)) |>
  select(all_of(core_site_fields), habitat)
```

#### Summary

```{r}
birds_sites_sf <- birds_sites |>
  filter(!is.na(longitude)) |> 
  sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  distinct(ps_site_id, region, habitat, geometry)

mapview::mapview(birds_sites_sf,
                 zcol = "habitat",
                 legend = TRUE,
                 col.regions = c("coastal" = "#B9E6FF", "inland" = "#F87575"),
                 map.types = "Esri.WorldImagery",
                 layer.name = "Habitat",
                 popup = leafpop::popupTable(pbruvs_sites_sf, zcol = c("ps_site_id", "region")))
```

### Submersible

#### QAQC

```{r}
sub_raw <- read_xlsx(file.path(exp_path, "data/primary/raw/sub/PNG_2024_sub_metadata.xlsx"),
                        sheet = "metadata") |>
  clean_names()

sub_sites <- sub_raw |> 
  transmute(### Core fields ----
            exp_id = "PNG_2024",
            survey_type = "sub",
            leg = "Leg 2",
            ps_site_id = str_replace(str_replace(ps_station_id, "(\\d+)$", \(x) str_pad(x, 3, pad = "0")), "PNG", "PNG_2024"),
            location = str_to_title(location),
            sublocation = str_to_title(sublocation),
            date = dmy(date_dd_mm_yyyy),
            time = as_hms(start_time_local),
            latitude = start_lat_surface,
            longitude = start_lon_surface,
            lead = observer_1,  # assuming primary observer =  lead
            notes = notes,
            ### Method-specific fields (sub) ----
            sub_name = "Argonauta",
            dive_number = dive_num,
            depth_max_m = as.numeric(depth_max_m),
            duration = as_hms(dive_duration),
            temp_max_depth_c = as.numeric(temp_max_depth_c),
            observer_1 = observer_1,
            observer2 = observer_2,
            pilot = pilot,
            dive_type = dive_type,
            collection = FALSE,
            transect = if_else(transect == "Y", TRUE, FALSE),
            edna = if_else(edna == "Y", TRUE, FALSE),
            # Waypoints
            ## Descent 
            time_descent = as_hms(start_time_local),
            lat_descent = start_lat_surface,
            lon_descent = start_lon_surface,
            ## Bottom
            time_on_bottom = NA,
            lat_on_bottom = NA_real_,
            lon_on_bottom = NA_real_,
            ## Off bottom
            time_off_bottom = NA,
            lat_off_bottom = NA_real_,
            lon_off_bottom = NA_real_,
            ## Surface
            time_surface = NA,
            lat_surface = NA_real_,
            lon_surface = NA_real_)

sub_sites <- sub_sites |> 
  left_join(region_lookup, by = "location") |>
  rename(subregion = location, 
         locality = sublocation) |> 
  replace_na(list(region = "Western Islands")) 
```

#### Summary

```{r}
sub_sites_sf <- sub_sites |>
  filter(!is.na(longitude)) |> 
  sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  distinct(ps_site_id, region,  geometry)

mapview::mapview(sub_sites_sf,
                 zcol = "region",
                 legend = TRUE,
                 col.regions = regions_pal,
                 map.types = "Esri.WorldImagery",
                 layer.name = "Region",
                 popup = leafpop::popupTable(pbruvs_sites_sf, zcol = c("ps_site_id", "region")))
```


### Dropcams

#### QAQC

```{r warning = FALSE}
dscm_raw <- read_xlsx(file.path(exp_path, "data/primary/raw/dscm/PNG_2024_dscm_fieldbook.xlsx"),
                        sheet = "meta") |>
  clean_names()

dscm_clean <- dscm_raw |> 
  transmute(### Core fields ----
            exp_id = "PNG_2024",
            survey_type = "dscm",
            leg = leg, 
            ps_site_id = str_replace(str_replace(ps_station_id, "(\\d+)$", \(x) str_pad(x, 3, pad = "0")), "PNG", "PNG_2024"),
            location = str_to_title(location),
            sublocation = str_to_title(location),
            date = ymd(date_in),
            time = as_hms(time_in),
            latitude = lat_in,
            longitude = lon_in,
            lead = team_lead,
            notes = notes,
            # Method fields
            extech_exp_id = extech_exp_id,
            dscm_id = dscm_id,
            mission_duration = as_hms(mission_duration_hrs),
            recording_duration = as_hms(recording_time_hrs),
            bottom_depth_m = round(as.numeric(depth_m),2),
            bottom_temp_c = round(as.numeric(bottom_temp_c),2),
            bottom_type = bottom_type,
            bait_type = bait_type,
            bait_kg = as.numeric(bait_kg),
            bait_can = bait_can,
            ballast_type = ballast_type,
            ballast_kg = as.numeric(bait_kg),
            gtr_type = gtr_type,
            iridium_external_id = iridium_external_id,
            highlights  =  highlights,
            # Waypoints
            time_deploy = as_hms(time_in),
            lat_deploy = lat_in,
            lon_deploy = lon_in,
            time_recovery = as_hms(time_out),
            lat_recovery = lat_out,
            lon_recovery = lon_out)

dscm_clean <- dscm_clean |> 
  left_join(region_lookup, by = "location") |>
  rename(subregion = location, 
         locality = sublocation) |> 
  mutate(region = case_when(subregion %in% c("Emanaus") ~ "Murat", 
                            subregion %in% c("Lonvongai") ~ "Lovongai", 
                            TRUE ~ region)) 
  
dscm_sites <- dscm_clean |> 
  select(all_of(core_site_fields))
```

#### Summary

```{r}
dscm_sites_sf <- dscm_sites |>
  filter(!is.na(longitude)) |> 
  sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  distinct(ps_site_id, region,  geometry)

mapview::mapview(dscm_sites_sf,
                 zcol = "region",
                 legend = TRUE,
                 col.regions = regions_pal,
                 map.types = "Esri.WorldImagery",
                 layer.name = "Region",
                 popup = leafpop::popupTable(pbruvs_sites_sf, zcol = c("ps_site_id", "region")))
```

## Merge and QAQC

```{r}
exp_sites <- bind_rows(uvs_sites, 
                       edna_sites, 
                       sbruvs_sites, 
                       pbruvs_sites, 
                       birds_sites, 
                       sub_sites, 
                       dscm_sites) 

# Check for duplicates
dupes <- exp_sites |> 
  janitor::get_dupes(ps_site_id)

if (nrow(dupes) > 0) {
  cli::cli_warn("⚠️ {nrow(dupes)} duplicate site(s) found.")
} else {
  cli::cli_alert_success("✅ No duplicate sites found.")
}

# Date and time range
min(exp_sites$date)
max(exp_sites$date)

# Locations

subregion_fixes <- c("Metenmin" = "Metemin",
                    "Metovai" = "Metevoi",
                    "Tsoi" = "Tsoilik",
                    "Tsolik" = "Tsoilik")

exp_sites <- exp_sites |>
  mutate(subregion = recode(subregion, !!!subregion_fixes))

exp_sites |> 
  group_by(region, subregion, survey_type) |> 
  summarize(sites = n()) |> 
  ungroup() |> 
  pivot_wider(values_from = sites, names_from = survey_type) |> 
  gt(groupname_col = "region") |> 
  tab_header(title = "Number of sites by region and subregion") |> 
  tab_options(row_group.as_column = T,
              table.font.size = "small",
              table.border.top.style = "solid",
              table.border.top.width = px(2),
              table.border.top.color = "black",
              heading.align = "center",
              column_labels.font.weight = "bold") |> 
  sub_missing(missing_text = "--")
```

## Coverage

```{r}
library(gt)

exp_sites |> 
  group_by(leg, region, survey_type) |> 
  summarize(n_sites = n_distinct(ps_site_id), .groups = "drop") |> 
  pivot_wider(names_from = survey_type, values_from = n_sites, values_fill = 0) |> 
  select(leg, region, uvs, edna, contains("bruvs"), birds, sub, dscm, everything()) |>
  gt(groupname_col = "leg") |>
  tab_header(title = "Sampling effort by region and method") |> 
  tab_stubhead(label = "Leg") |> 
  tab_options(row_group.as_column = T,
              table.font.size = "small",
              table.border.top.style = "solid",
              table.border.top.width = px(2),
              table.border.top.color = "black",
              heading.align = "center",
              column_labels.font.weight = "bold")
```

```{r}
library(leaflet)
library(leaflet.extras)
library(dplyr)

# Optional: assign colors to methods
method_palette <- colorFactor(palette = "Set2", domain = unique(exp_sites$survey_type))

# Create map
leaflet_map <- leaflet(exp_sites) |>
  addProviderTiles(providers$Esri.WorldImagery) |> 
  addScaleBar(position = "bottomleft") |>
  addFullscreenControl()

# Add a marker group for each method
methods <- unique(exp_sites$survey_type)

for (m in methods) {
  
  method_data <- exp_sites |> 
    filter(survey_type == m) |> 
    filter(!is.na(latitude) & !is.na(longitude))
  
  leaflet_map <- leaflet_map |>
    addCircleMarkers(data = method_data,
                     ~longitude, ~latitude,
                     color = ~method_palette(survey_type),
                     radius = 3,
                     stroke = TRUE,
                     fillOpacity = 0.8,
                     group = m,
                     label = ~ps_site_id,
                     popup = ~paste("<b>Site ID:</b>", ps_site_id, "<br>",
                                    "<b>Method:</b>", survey_type, "<br>",
                                    "<b>Date:</b>", as.character(date), "<br>",
                                    "<b>Location:</b>", region, "<br>",
                                    "<b>Habitat:</b>", habitat   ))
  }

# Add layer toggle controls

leaflet_map <- leaflet_map |>
  addLayersControl(overlayGroups = methods,
                   options = layersControlOptions(collapsed = FALSE)) |> 
  addSearchFeatures(targetGroups = unique(exp_sites$survey_type),
                    options = searchFeaturesOptions(zoom = 12,
                                                    openPopup = TRUE,
                                                    firstTipSubmit = TRUE,
                                                    autoCollapse = TRUE,
                                                    hideMarkerOnCollapse = TRUE))

leaflet_map
```

