---
title: "eda_SST_satellite"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

## Load Packages

```{r}

rm(list = ls())

pkgTest <- function(x)
{
  if (!require(x,character.only = TRUE))
  {
    install.packages(x,dep=TRUE,repos='http://cran.us.r-project.org')
    if(!require(x,character.only = TRUE)) stop(x, " :Package not found")
  }
}


# Create list of required packages
list.of.packages <- c("ggplot2", "dplyr", "scales","RColorBrewer", "readxl","tidyverse", "stringr","stringi")
 

# Create list of installed packages
pkges = installed.packages()[,"Package"]

# Install and load all required pkgs
for (pk in list.of.packages) {
  pkgTest(pk)
}

```

## Data Setup

Loading the RData file that has the SST time series data of the mean monthly values for PNG for each site from 2014-2025 along with the maximum monthly mean (MMM) values by region and subregion which were calculated from the Coral Reef Watch (CRW) climatological range of 1985-2012.

```{r}

load("~/GitHub/PNG_2024/PNG_satellite_SST.RData")



```

### Maximum Monthly Mean (MMM) and Bleaching Threshold graph

Using the SST values, grouping sites into regions and subregions, calculating means and graphing results, adding the MMM and threshold lines.

#### Region

```{r}

PNG_month_region <- PNG_SST_monthly_means %>%
  mutate(date = substr(date, 0, 10)) %>% 
  group_by(date, region) %>%
  dplyr::summarize(
    count = n(),              # number of observations
    SSTmean = mean(SST, na.rm = TRUE),    # mean value
    SSTmedian = median(SST, na.rm = TRUE),      # sum value
    SSTmax = max(SST, na.rm = TRUE),      # maximum valuex  
    SSTmin = min(SST, na.rm = TRUE),      # minimum value
  ) %>%
  mutate(date2 = as.Date(date)) %>% 
  mutate(year = substr(date,0,4)) %>% 
  mutate(yrmonth = substr(date,0,7)) %>% 
  ungroup() 

PNG_month_region<-PNG_month_region %>% 
  left_join(PNG_region_thresholds, by = "region")

ggplot(PNG_month_region,aes(x=date2, y=SSTmean, color = region))+
    geom_point(size = 1)+
  ylim(26,31)+
#  scale_x_date(date_breaks = "1 month", date_labels = "%m")+
  geom_line(aes(group=region))+
  ylab("Mean Monthly \nSea Surface Temperature")+
  scale_x_date(breaks = date_breaks("2 year"), date_labels = "%Y")+
  xlab("")+
 scale_color_manual(values=c("navy","orange", "dodgerblue","cyan", "green4", "lightblue"))+

  facet_wrap(~region)+
    geom_hline(aes(yintercept = mean, linetype = "Maximum Monthly Mean (MMM)"), size= .25, color = "tomato4")+
  geom_hline(aes(yintercept = Kleypass_threshold, linetype = "Bleaching Threshold"),
  size= .25, color = "tomato4")+
  labs(linetype = NULL) +
          theme_bw()+
   theme(axis.text.x = element_text(angle = 0, size = 8),
         legend.key = element_blank())


```

#### SubRegion

```{r}

PNG_month_subregion <- PNG_SST_monthly_means %>%
  mutate(date = substr(date, 0, 10)) %>% 
  group_by(date,region, subregion) %>%
  dplyr::summarize(
    count = n(),              # number of observations
    SSTmean = mean(SST, na.rm = TRUE),    # mean value
    SSTmedian = median(SST, na.rm = TRUE),      # sum value
    SSTmax = max(SST, na.rm = TRUE),      # maximum valuex  
    SSTmin = min(SST, na.rm = TRUE),      # minimum value
  ) %>%
  mutate(date2 = as.Date(date)) %>% 
  mutate(year = substr(date,0,4)) %>% 
  mutate(yrmonth = substr(date,0,7)) %>% 
  ungroup() #%>%   filter(year == "2024")

PNG_month_subregion<-na.omit(PNG_month_subregion)


PNG_month_subregion<-PNG_month_subregion %>% 
  left_join(PNG_subregion_thresholds, by = "subregion")

ggplot(PNG_month_subregion,aes(x=date2, y=SSTmean, color = subregion))+
    geom_point(size = 1)+
  ylim(29,31)+
  geom_line(aes(group=subregion))+
  ylab("Mean Monthly \nSea Surface Temperature")+
  scale_x_date(breaks = date_breaks("2 year"), date_labels = "%Y")+
  xlab("")+
   facet_wrap(~subregion)+
    geom_hline(aes(yintercept = mean, linetype = "Maximum Monthly Mean (MMM)"), size= .25, color = "tomato4")+
  geom_hline(aes(yintercept = Kleypass_threshold, linetype = "Bleaching Threshold"),
  size= .25, color = "tomato4")+
  labs(linetype = NULL) +
          theme_bw()+
   theme(axis.text.x = element_text(angle =90, vjust = 0.5,size = 8),
         legend.key = element_blank())
```

### Daily

Selecting the worse sites that showed bleaching or death from Lovongai and Manus

```{r}
lov_metric<-PNG_SST_daily_metrics %>% 
  filter(region == "Lovongai") %>% 
  filter(ps_site_id %in% c("PNG_2024_uvs_034","PNG_2024_uvs_035","PNG_2024_uvs_038", "PNG_2024_uvs_033", "PNG_2024_uvs_023","PNG_2024_uvs_040")) %>% 
   mutate(date2 = as.Date(date)) %>% 
  mutate(monthday= str_sub(date2, 6,-1)) %>% 
  left_join(PNG_subregion_thresholds, by = "subregion")

manus_metric<-PNG_SST_daily_metrics %>% 
  filter(region == "Manus") %>% 
  filter(ps_site_id %in% c("PNG_2024_uvs_046","PNG_2024_uvs_051","PNG_2024_uvs_052", "PNG_2024_uvs_053", "PNG_2024_uvs_058","PNG_2024_uvs_059")) %>% 
   mutate(date2 = as.Date(date)) %>% 
  mutate(monthday= str_sub(date2, 6,-1)) %>% 
  left_join(PNG_subregion_thresholds, by = "subregion")

```

#### SST

```{r}

ggplot(lov_metric,aes(x=date2, y=SST, color = ps_site_id))+
    geom_point(size = 1)+
  ylim(26,31)+
#  scale_x_date(date_breaks = "1 month", date_labels = "%m")+
  geom_line(aes(group=subregion))+
  ylab("Mean Monthly \nSea Surface Temperature")+
 scale_color_manual(values=c("navy","orange", "thistle3","steelblue3", "tan","yellow3" ))+
  xlab("")+
  scale_x_date(breaks = date_breaks("1 month"), date_labels = "%m-%d")+
  facet_wrap(~subregion)+
    geom_hline(aes(yintercept = mean, linetype = "Maximum Monthly Mean (MMM)"), size= .25, color = "tomato4")+
  geom_hline(aes(yintercept = Kleypass_threshold, linetype = "Bleaching Threshold"),
  size= .25, color = "tomato4")+
  labs(linetype = NULL) +
          theme_bw()+
   theme(axis.text.x = element_text(angle =90, hjust = 1, vjust = 0.5, size = 8))



```

#### DHW

```{r}

ggplot(lov_metric,aes(x=date2, y=DHW, color = ps_site_id))+
    geom_point(size = 1)+
  geom_line(aes(group=subregion))+
  ylab("DHW")+
  xlab("")+
   scale_color_manual(values=c("navy","orange", "thistle3","steelblue3", "tan","yellow3" ))+
  scale_x_date(breaks = date_breaks("1 month"), date_labels = "%m-%d")+
  facet_wrap(~subregion)+
          theme_bw()+
   theme(axis.text.x = element_text(angle =90, hjust = 1, vjust = 0.5, size = 8))



```

#### Anomaly

```{r}

ggplot(lov_metric,aes(x=date2, y=Anomaly, color = ps_site_id))+
    geom_point(size = 1)+
  geom_line(aes(group=subregion))+
  ylab("Temperature Anomalies")+
  xlab("")+
 scale_color_manual(values=c("navy","orange", "dodgerblue","cyan", "magenta","yellow3" ))+
  scale_x_date(breaks = date_breaks("1 month"), date_labels = "%m")+
  facet_wrap(~subregion)+
          theme_bw()+
   theme(axis.text.x = element_text(angle = 0, size = 8),
         legend.key = element_blank())



```

#### Bleaching Alert Level

```{r}

ggplot(manus_metric,aes(x=date2, y=BAA, color = ps_site_id))+
    geom_point(size = 1)+
  geom_line(aes(group=subregion))+
  ylab("Bleaching Alert Level")+
  xlab("")+
 scale_color_manual(values=c("navy","orange", "thistle3","steelblue3", "tan","yellow3" ))+
  scale_x_date(breaks = date_breaks("1 month"), date_labels = "%m-%d")+
  facet_wrap(~subregion)+
          theme_bw()+
   theme(axis.text.x = element_text(angle =90, hjust = 1, vjust = 0.5, size = 8))



```

#### HotSpot

```{r}

ggplot(lov_metric,aes(x=date2, y=HotSpot, color = ps_site_id))+
    geom_point(size = 1)+
  geom_line(aes(group=subregion))+
  ylab("HotSpot")+
  xlab("")+
 scale_color_manual(values=c("navy","orange", "dodgerblue","cyan", "magenta","yellow3" ))+
  scale_x_date(breaks = date_breaks("1 month"), date_labels = "%m-%d")+
  facet_wrap(~subregion)+
          theme_bw()+
   theme(axis.text.x = element_text(angle =90, hjust = 1, vjust = 0.5, size = 8))


```
